A note on observation processes in epidemic models

### 2019-11-25 ms/ms.tex
edits

Note that incidence (the number of \emph{newly} infected individuals) is different from 
prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics};
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. 
While they are relatively
uncommon, some models do not make a clear distinction between prevalence and 
incidence 
\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.

Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how 
assumptions about the underlying observation processes affect parameter estimates
of the SIR model. We show that making wrong assumptions about the timing of 
incidence reports can lead to biased parameter estimates and overly narrow 
confidence intervals.

@@

The Susceptible-Infected-Removed (SIR) model describes how a disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
and $N = S + I + R$ is the total population size. 
We define incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected cases to be reported some time after infection;
the number of reported cases during a time period defines the \emph{observed} incidence.

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection, $i_1(t)$, can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
\end{equation}
Equivalently, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model reporting error using a negative binomial distribution with a
mean of either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
similarly, we will refer to epidemic time series generated from these
negative binomial models with two different means ($\rho i_1(t)$ and $\rho i_2(t)$)
as infection time series and recovery time series.

In this study, we focus on estimating 5 parameters: the basic reproductive
number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
reporting rate $\rho$, an overdispersion parameter $\theta$, and the initial
proportion of the infected individuals $i_0$. The initial proportion of
susceptible individuals is assumed to be $1 - i_0$. The total population
size $N$ is assumed to be known.

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
A deterministic SIR model is simulated using the following parameters: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
}
\end{figure}

Figure 1 compares the deterministic dynamics of incidence curves,
$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$. A lag in reporting time delays
the observed epidemic peak timing and slightly reduces the size of the peak. However, 
these differences in the reporting time have little effect on the overall shape 
of the epidemic curve. In the presence of observation and process error, we
do not expect to be able to distinguish between the two reporting processes
based on the time series alone (and we rarely know \emph{a priori} the delay distribution between when an individual is infected and when that case is reported).
Therefore, one might naively expect assumptions about the timing of case reporting to have negligible effect on inference.

In order to understand how assumptions about the timing of case reporting affect 
parameter estimates of the SIR model, we simulate observed epidemic time series
(infection time series and recovery time series) 
100 times and fit both infection and recovery models to each 
time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
and the coverage of our confidence intervals, defined as the proportion of confidence intervals that
contain the true value (95\% confidence interval is expected to contain the true value
95\% of the time by definition). Figure 2 summarizes the results.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias (in opposite directions) 
and a much lower coverage (Figure 2B).
\swp{Cite elderd}

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, epidemic models often make 
simple assumptions about how cases are reported. 
Here, we used a deterministic SIR model to show that 
incorrect assumptions about observational processes in epidemic models
can bias estimates of the basic reproduction number and lead to overly narrow
confidence intervals.
\bmb{repetitive}

@@

There is some empirical evidence that delays in case notification alters the
shape of an epidemic curve and affects our understanding of disease transmission:
(1) Postal delays in measles notification during holidays lead to sudden
epidemic peaks after holidays \citep{fine1982measles}; 
(2) Delays in reports of the foot-and-mouth infection affects the effectiveness
of slaughter policies on epidemic intervention;
and (3) Date of wills written and date of wills probated during
the plague epidemics in London give completely different looking epidemic curves
due to extremely long delays (mean of 2.9 years) between the two \citep{bushby2019wills}.
Ignoring such delays in reporting may affect conclusions about the underlying 
transmission mechanisms.

Here, we assumed that the underlying transmission process is deterministic; this 
assumes that all error can be explained by observation errors alone. 
We chose to study a deterministic model SIR for computational efficiency; 
we do not recommend using deterministic models for real outbreak analyses.
Ignoring process errors (i.e., stochasticity in the transmission process) can lead to
overly confident results \citep{king2015avoidable}. Using stochastic models may
even give better coverage probabilities when wrong observation models are used. 
Nonetheless, misspecifying the observation model can still affect conclusions from 
stochastic models and introduce bias.

>>>>

Note that incidence (the number of \emph{newly} infected individuals) is different from 
prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics}:
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not.
While they are relatively
uncommon, some models do not make a clear distinction between prevalence and 
incidence 
\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.

Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how
assumptions about the underlying observation processes affect parameter estimates
of the SIR model. We show that making incorrect assumptions about the timing of 
incidence reports can lead to biased parameter estimates and overly narrow 
confidence intervals.

@@

The Susceptible-Infected-Removed (SIR) model describes how a disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
and $N = S + I + R$ is the total population size. 
We define \emph{true} incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected cases to be reported some time after infection;
the number of reported cases during a time period defines the \emph{observed} incidence

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection, $i_1(t)$, can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
\end{equation}
Equivalently, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
the observed incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model observation error using a negative binomial distribution with a
mean of either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
similarly, we will refer to epidemic time series generated from these
negative binomial models with two different means ($\rho i_1(t)$ and $\rho i_2(t)$)
as infection time series and recovery time series.

In this study, we focus on estimating 5 parameters: the basic reproductive
number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
reporting rate $\rho$, the overdispersion parameter $\theta$, and the initial
proportion of the infected individuals $i_0$. The initial proportion of
susceptible individuals is assumed to be $1 - i_0$. The total population
size $N$ is assumed to be known.

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/example.pdf}
\caption{
\textbf{A comparison of incidence measured at two different time points in infection.}
(A) 
A deterministic simulation of the SIR model using the following parameters: 
$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
and $\Delta t = 0.1$ time units.
(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to lags in reporting time. 
Black circles: simulation results of the deterministic SIR model.
Black lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
}
\end{figure}

Figure 1A compares the deterministic dynamics of two incidence curves,
$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$ for $\mathcal R_0 = 2$. 
A lag in reporting time delays
the timing of the observed epidemic peak and reduces the size of that peak.
As $\mathcal R_0$ increases, the difference in the size of the peaks increases (Figure 1B)
but the difference in the timing of the peaks decreases (Figure 1C).
For example, when $\mathcal R_0 = 5$, such delay in the reporting of new cases can reduce the size of the observed epidemic peak by almost 50\%.

For small values of $\mathcal R_0$, 
these differences in the reporting time have little effect on the overall shape 
of the epidemic curve. In the presence of observation and process error, we
do not expect to be able to distinguish between the two reporting processes
based on the time series alone (and we rarely know \emph{a priori} the delay distribution between when an individual is infected and when that case is reported).
Therefore, one might naively expect assumptions about the timing of case reporting to have negligible effect on inference.

In order to understand how assumptions about the timing of case reporting affect 
parameter estimates of the SIR model, we simulate observed epidemic time series
(infection time series and recovery time series) 
100 times with $\mathcal R_0 = 2$ and fit both infection and recovery models to each 
time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
and the coverage of our confidence intervals, defined as the proportion of confidence intervals that
contain the true value (95\% confidence interval is expected to contain the true value
95\% of the time by definition). Figure 2 summarizes the results.

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the true value of the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting the
incorrect model results in a clearer bias (in opposite directions)
and a much lower coverage (cf. \cite{elderd2006uncertainty}).

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, epidemic models often make 
simple assumptions about how cases are reported. 
We used a deterministic SIR model to show that
assumptions about observation processes can
affect the estimates of the basic reproduction number and their associated confidence intervals.

@@

Here, we assumed that the underlying transmission process is deterministic; this 
assumes that all error can be explained by observation errors alone. 
We chose to study a deterministic model SIR for computational efficiency; 
we do not recommend using deterministic models for real outbreak analyses.
Ignoring process errors (i.e., stochasticity in the transmission process) can lead to
overly confident results \citep{king2015avoidable}. Using stochastic models may
give better coverage probabilities even  when wrong observation models are used.
Nonetheless, misspecifying the observation model can still affect conclusions from 
stochastic models and introduce bias.
